The Effects of the NBA COVID Bubble on the NBA Playoffs

### 2020-11-04 Manuscript/NBABubbleAnalysis.Rmd
abstract

---
title: "Analyzing the Effects of the NBA COVID Bubble on the NBA Playoffs"
subtitle: "A Case Study for Home-Court Advantage"
author: |
  | Michael Price
  | michael.price@uconn.edu
  | Department of Statistics, University of Connecticut
  | Dr. Jun Yan
  | jun.yan@uconn.edu
  | Department of Statistics, University of Connecticut

date: "`r format(Sys.time(), '%d, %B, %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: citations.bib
biblio-style: asa
output: bookdown::pdf_document2
header-includes:
   - \usepackage{float}
toc: False
---
Abstract: 
The 2020 NBA playoffs were played inside of a bubble in Disney World because of
the COVID-19 pandemic. This meant that there would be no fans in attendance, 
games played on neutral courts and no traveling for teams which in theory 
removes home-court advantage from the games. This idea was a hot topic of 
discussion as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played a huge role in 
NBA playoff series outcomes. Thus, to study the effects of the bubble, and 
home-court advantage, the 2020 season is compared to the 2017, 2018, and 2019
playoffs. While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins games so scoring will be the primary focus 
of this study. The specific measures of interest are team scoring totals and 
team shooting percentage on two pointers, three pointers, and free throws. 
Comparing these measures for home teams and away teams in 2020 vs. 2017-2019
shows how the 2020 playoffs favored away teams more than normal, particularly
with two point shooting and total scoring.

```{r setup, echo = FALSE, message =  FALSE, warning = FALSE}
library(BSDA)
library(ggplot2)
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, dpi = 300,
out.width = "90%", fig.align = "center", dev = "pdf", fig.pos = "H")
```
\newpage
# I. Introduction  
Home-court advantage is often discussed in sports circles as a contributing 
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge from playing at their home-court, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a 
bubble brought a great deal of concern for fans, teams, journalists, and others. 
Steve Aschburner discusses details about anticipated effects in his article 
“How Orlando's Neutral Site Will Impact Teams, Referees and Games” @Aschburner. 
Aschburner shares concerns from former players, coaches and other experts about
the potential effects of removing home-court advantage. Aschburner notes that the
NBA did make attempts to recreate the effects by putting the "home” team logo on
the court and allowing the "home" team to play crowd noise and music, but most 
people doubted these small attempts would recreate a true playoff atmosphere. 
During the 2020 NBA playoffs, home teams only won about 48.2% of the games. This
is lower than normal, which Aschburner claims usually floats around 60%. This 
shift in the home team winning percentage surely indicates the opportunity for
thorough investigation.

>>>>

---
title: "The Effects of the NBA COVID Bubble on the NBA Playoffs"
subtitle: "A Case Study for Home-Court Advantage"
author: |
  | Michael Price and Jun Yan
  | michael.price@uconn.edu
  | Department of Statistics, University of Connecticut
date: "`r format(Sys.time(), '%d, %B, %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: citations.bib
biblio-style: asa
output: bookdown::pdf_document2
header-includes:
   - \usepackage{float}
   - \usepackage[margin=1in]{geometry}
toc: False
---
Abstract: 
The 2020 NBA playoffs were played inside of a bubble in Disney World because of
the COVID-19 pandemic. This meant that there would be no fans in attendance, 
games played on neutral courts and no traveling for teams, which in theory 
removes home-court advantage from the games. This setting has attracted a lot of
discussion as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played a huge role in 
NBA playoff series outcomes. The 2020 playoff provided a unique
opportunituy to study the effects of the bubble, and home-court
advantage by comparing the 2020 season with the seasons in the past
While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins games so scoring is the primary focus 
of this study. The specific measures of interest are team scoring totals and 
team shooting percentage on two pointers, three pointers, and free throws. 
Comparing these measures for home teams and away teams in 2020 vs. 2017-2019
shows that the 2020 playoffs favored away teams more than normal, particularly
with two point shooting and total scoring.

```{r setup, echo = FALSE, message =  FALSE, warning = FALSE}
library(BSDA)
library(ggplot2)
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, dpi = 300,
out.width = "90%", fig.align = "center", dev = "pdf", fig.pos = "H")
```

# Introduction  
Home-court advantage is often discussed in sports circles as a contributing 
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge from playing at their home-court, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a 
bubble brought a great deal of concern for fans, teams, journalists, and others. 
Steve Aschburner discusses details about anticipated effects in his article 
“How Orlando's Neutral Site Will Impact Teams, Referees and Games” @Aschburner. 
Aschburner shares concerns from former players, coaches and other experts about
the potential effects of removing home-court advantage. Aschburner notes that the
NBA did make attempts to recreate the effects by putting the "home” team logo on
the court and allowing the "home" team to play crowd noise and music, but most 
people doubted these small attempts would recreate a true playoff atmosphere. 
During the 2020 NBA playoffs, home teams only won about 48.2% of the games. This
is lower than normal, which Aschburner claims usually floats around 60%. This 
shift in the home team winning percentage surely indicates the opportunity for
thorough investigation.
